On The Subspace Learning for Network Attack Detection

### 2019-06-09 docs/thesis/chapters/4_m_rpca/m_rpca.tex
results of robpca

In this paper, scalars are denoted by italic letters (\emph{a, b, A, B, $\alpha$, $\beta$}), vectors by lowercase bold letters (\textbf{a, b}), matrices by uppercase bold letters (\textbf{A, B}), and $a_i,_j$ denotes the (\emph{i, j}) elements of the matrix \textbf{A}. The superscripts \textsuperscript{T} and \textsuperscript{-1} are used for matrix transposition and matrix inversion, respectively. The Frobenius norm is denoted as $\left\| \mathord{\cdot} \right\|_F$, while $\left\| \mathord{\cdot} \right\|_*$ denotes the nuclear norm of a matrix and $\left\| \mathord{\cdot} \right\|_1$ means the sum of the absolute values of matrix entries. We also define the operator $\langle \mathord{\cdot} \rangle$, that is the standard trace inner product, and $[ \mathord{\cdot} ]^n$, which denotes the $n$ largest values of a vector.

@@

In order to evaluate the null hypothesis $H_{0rob.sub}$, which assumes that the robust subspace learning does not improves the anomaly detection from imbalanced and skewed, we create two synthetic data sets characterized by skewed and heavy tailed distributions. We selected Pareto and Lognormal distributions to evaluate the anomaly detection capacity of our proposal, considering that these distributions have been adopted to represent network traffic in the internet and data centers \cite{benson2010network,leon2017probability}.

We adopt a contamination rate of 33\% to reproduce the imbalanced data of anomaly detection problems, and select the gaussian distribution for the addictive anomalies into Pareto and Lognormal distributions. Therefore, we generate $\boldsymbol{Y}_p$ and $\boldsymbol{Y}_l$ with 2,400 samples and two features, where 793 samples are gaussian anomalies, also referred as noise or contamination, while 1,607 samples are legitimate signals. The data set $\boldsymbol{Y}_p$ is composed by random Pareto distribution with random gaussian noise, while $\boldsymbol{Y}_l$ is composed by random Lognormal distribution with random gaussian noise.

The Figure \ref{fig:4.03} shows the histogram and the PDF of the Pareto distribution for two features of $\boldsymbol{Y}_p$.

\begin{figure}[!htb]
	\centering
	\subfigure[Feature 1]{%
		\includegraphics[width=0.48\textwidth]{figures/pareto1.png}
		\label{fig:4.01}
	}
	\subfigure[Feature 2]{%
		\includegraphics[width=0.48\textwidth]{figures/pareto2.png}
		\label{fig:4.02}
	}
	\caption[Pareto Distribution]{Pareto Distribution}
	\label{fig:4.03}
\end{figure}

The Figure \ref{fig:4.06} shows the histogram and the PDF of the Lognormal distribution for two features of $\boldsymbol{Y}_l$.

\begin{figure}[!htb]
	\centering
	\subfigure[Feature 1]{
		\includegraphics[width=0.48\textwidth]{figures/lognormal3.png}
		\label{fig:4.04}
	}
	\subfigure[Feature 2]{
		\includegraphics[width=0.48\textwidth]{figures/lognormal4.png}
		\label{fig:4.05}
	}
	\caption[Lognormal Distribution]{Lognormal Distribution}
	\label{fig:4.06}
\end{figure}

The Pareto distribution contaminated by gaussian noise is denoted as $\boldsymbol{Y}_p^c$, while $\boldsymbol{Y}_l^c$ denotes the Lognormal distribution contaminated by gaussian noise. Taking into account that semi-supervised algorithms requires training from normal data, we also generate $\boldsymbol{Y}_p^t$, while $\boldsymbol{Y}_l^t$, which are data sets of Pareto and Lognormal distributions for training, without contamination and composed by 1,607 samples and two features.

@@

The CTU-13 is a data set of botnet traffic that was captured in the Czech Technical University \cite{garcia2014empirical}, by means of a testbed and execution of malwares in a real network. The CTU-13 data set contains 13 scenarios with network flows of botnet malwares, that are: neris, rbot, virut, menti, sogou, nsys.ay and murlo. The botnet traffic is also classified as attack or command and control (C\&C), while the legitimate flows can also be classified as normal or background. It is important to note that the CTU-13 data set can also be formulated as a superposition of signal, artefact and noise, which refer to background, normal a botnet traffic, respectively.

@@

We refer to Garcia \cite{garcia2014identifying} and Garcia \emph{et al.} \cite{garcia2014empirical} for a detailed description of the performed attacks and C\&C flows, also including the topology of the adopted testbed, rules for classifying normal flows, and an analysis of behaviors or patterns of the malware's traffic.

For all the scenarios, the authors of the CTU-13 data set convert the captured pcap files to NetFlows and release the processed flows. The data set contains ground-truth labels for flows as follows: flows from or to the infected machines are labeled as “botnet”; flows from or to well known and controlled machines are labeled as “normal”; all other flows are labeled as “background.”

Table \ref{tab:4.01} presents an overview organized by scenario and shows the malwares used for botnet attacks, the types of attacks and C\&C types, the total number of flows, the number of malicious flows, which includes flows of C\&C and attacks, and finally shows the number of normal flows.

@@

The CTU-13 data set originally contains the following features:

\begin{itemize}
	\item Start Time;
	\item End Time;
	\item Duration;
	\item Source IP Address;
	\item Source Port;
	\item Direction;
	\item Destination IP Address;
	\item Destination Port;
	\item State;
	\item Type of service from source to destination;
	\item Type of service from destination to source;
	\item Total of Packets;
	\item Total of Bytes;
\end{itemize}

Our analysis of the available features leads to discard some features, considering that highly correlated features can bias or not improve the model, and that source or destination IP addresses can insert some false bias into learning models, since they can be changed by IP spoofing or other techniques.

TODO: talk about the feature distribution, that arent similar to pareto or lognormal, but it is not normal too

An Exploratory Data Analysis (EDA) reveals that distributions of some features are skewed and present an overlapping between normal and anomalous flows, as can be seen in  \ref{fig:4.01}, that presents the distribution of transaction state of scenario 16, and exemplifies the data skewness, the overlapping and the long tail distribution of normal data. 
\begin{figure}[h!]
     \centering
     \includegraphics[width=10cm]{figures/raw_distplot_capture20110816_State.png}
     \caption{Distribution of flow duration for scenario 12.}
     \label{fig:4.01}
\end{figure}

Due to the number of available features and the class overlapping, we performed an correlation analysis and cross validation in order to identify the best set of features for network attack detection. Therefore, we adopt the following features: State, Type of service from source to destination, Type of service from destination to source, destination port, source port, and the total number of packets.

The full data set and scenarios can be denoted as $\boldsymbol{X} = \{\boldsymbol{X}_{10}, \boldsymbol{X}_{11}, \ldots , \boldsymbol{X}_{18-2}, \boldsymbol{X}_{19}\}$, in accordance to IDs presented in Table \ref{tab:4.01}. Each contaminated scenario $\boldsymbol{X}_i$ is split into $\boldsymbol{X}_i^n$ containing 50\% of the the normal data, and into $\boldsymbol{X}_i^t$ that is composed by all anomalous flows and ne needed number of normal flows to have a testing data with 33\% of contamination.

@@

Robust subspace learning can be defined as the decomposition of a given data matrix $\textbf{X}$ into the sum of a low rank matrix $\textbf{L}$, whose column subspace gives the principal components, and a sparse matrix $\textbf{S}$, with outliers or noise. We propose to learn a robust subspace $\textbf{L}$ from the normal traffic $\textbf{X}^r$ for computing $\textbf{L}^r$, $\textbf{S}^r$ and the robust moments, i.e. the mean $\boldsymbol{\bar{x}}$, skewness $\boldsymbol{\bar{s}}$ and kurtosis $\boldsymbol{\bar{k}}$, in order to evaluate the distance $\boldsymbol{d}$ of new observations $\boldsymbol{X}^t$ to these moments and classify as anomalous the observations with the largest distances.

@@

Before ALM, some methods were proposed to solve that convex optimization problem, such as Iterative Thresholding (IT) and Accelerated Proximal Gradient (APG). However, according to \cite{lin2010augmented}, both approaches have scalability problems and require a large number of iterations to converge. The Augmented Lagrange Multiplier is proven to have a \emph{Q-linear} convergence speed and experimental results show that ALM is five times faster than APG, which in theory is sub-linear \cite{lin2010augmented}. Furthermore, ALM reaches more accurate results with less iterations.

@@

We propose to compute the mean $\boldsymbol{\bar{x}}$, the skewness $\boldsymbol{\bar{s}}$, the kurtosis $\boldsymbol{\bar{k}}$ and the covariance matrix $\boldsymbol{\bar{\Sigma}}$ from the robust subspace $\boldsymbol{L}^r$, after to minimize the Equation \ref{eq:4.06} for $\textbf{X}^r$. We also propose to compute the Mahalanobis Distance (MD) for detecting anomalies between new observations and the moments calculated from a robust subspace computed by RPCA.

@@

The contamination rate parameter is widely adopted for well established outlier detection algorithms \cite{zhao2019pyod}, and refers to the percentage of observations that are classified as anomalous. The contamination rate can be well known for some areas, or can be computed by cross-validation or can be assumed according to previous observations. In our proposal, the contamination define the rate of highest distances $\boldsymbol{d}$ that shall be classified as anomalous. Therefore, the resulting $\boldsymbol{\bar{t}}$ refers to observations classified as anomalous according to the contamiation $c$.

The Algorithm \ref{alg:4.01} presents how the m-RPCA can be algorithmically described.
\begin{algorithm}
	\label{alg:4.01}
	\SetAlgoLined
	\KwResult{$\boldsymbol{\bar{t}}_{\bar{\boldsymbol{x}}}$, $\boldsymbol{\bar{t}}_{\bar{\boldsymbol{s}}}$, $\boldsymbol{\bar{t}}_{\bar{\boldsymbol{k}}}$}
	Given $\boldsymbol{X}$ split into $\boldsymbol{X}^n$ and $\boldsymbol{X}^t$\;
	\While{not $\min_{L,S}\left \| \boldsymbol{L} \right \|_{*} + \lambda \left \| \boldsymbol{S} \right \|_{1}$ of $\boldsymbol{X}^n$}{
		$\boldsymbol{L}_{k+1} \in \operatorname*{argmin}_{\boldsymbol{L} \in \mathbb{R}^{m \times n}}\{l(\boldsymbol{L}, \boldsymbol{S}_{k}, \boldsymbol{Y}_{k})\}$\;
		$\boldsymbol{S}_{k+1} \in \operatorname*{argmin}_{\boldsymbol{S} \in \mathbb{R}^{m \times n}}\{l(\boldsymbol{L}_{k+1}, \boldsymbol{S}, \boldsymbol{Y}_{k})\}$\;
		$\boldsymbol{Y}_{k+1} = \boldsymbol{Y}_{k} + \mu(\boldsymbol{X} - \boldsymbol{L}_{k} - \boldsymbol{S}_{k})$\;
	}	
	$\bar{\boldsymbol{x}} = \displaystyle\frac{1}{h}\displaystyle\sum_{i\in \boldsymbol{h}} \boldsymbol{x}_i$, for $\boldsymbol{L}^n$\;
	$\boldsymbol{\bar{\Sigma}} = \displaystyle\frac{1}{h}\displaystyle\sum_{i\in \boldsymbol{h}} (\boldsymbol{x}_i - \bar{\boldsymbol{x}})(\boldsymbol{x}_i - \bar{\boldsymbol{x}})^\prime$, for $\boldsymbol{L}^n$\;
	$\boldsymbol{\bar{s}} = \frac{m_3}{m_2^{\frac{3}{2}}}$, for $\boldsymbol{L}^n$\;
	$\boldsymbol{\bar{k}} = \frac{m_4}{m_2^2}$, for $\boldsymbol{L}^n$\;
	$\boldsymbol{d}(\boldsymbol{x},\bar{\boldsymbol{x}}, \boldsymbol{\bar{\Sigma}}) = \sqrt{(\boldsymbol{x} - \bar{\boldsymbol{x}}) \boldsymbol{\bar{\Sigma}}^{-1}(\boldsymbol{x} - \bar{\boldsymbol{x}})^\prime}$\;
	$\boldsymbol{d}(\boldsymbol{x}, \bar{\boldsymbol{s}}, \boldsymbol{\bar{\Sigma}}) = \sqrt{(\boldsymbol{x} - \bar{\boldsymbol{s}}) \boldsymbol{\bar{\Sigma}}^{-1}(\boldsymbol{x} - \bar{\boldsymbol{s}})^\prime}$\;
	$\boldsymbol{d}(\boldsymbol{x}, \bar{\boldsymbol{k}}, \boldsymbol{\bar{\Sigma}}) = \sqrt{(\boldsymbol{x} - \bar{\boldsymbol{k}}) \boldsymbol{\bar{\Sigma}}^{-1}(\boldsymbol{x} - \bar{\boldsymbol{k}})^\prime}$\;
	$\boldsymbol{\bar{t}}_{\bar{\boldsymbol{x}}} = [\boldsymbol{d}(\boldsymbol{x}, \bar{\boldsymbol{x}}, \boldsymbol{\bar{\Sigma}})]^c$\;
	$\boldsymbol{\bar{t}}_{\bar{\boldsymbol{s}}} = [\boldsymbol{d}(\boldsymbol{x}, \bar{\boldsymbol{s}}, \boldsymbol{\bar{\Sigma}})]^c$\;
	$\boldsymbol{\bar{t}}_{\bar{\boldsymbol{k}}} = [\boldsymbol{d}(\boldsymbol{x}, \bar{\boldsymbol{k}}, \boldsymbol{\bar{\Sigma}})]^c$\;
	\caption{Moment Distances from Robust Subspace}
\end{algorithm}

The steps between 1 and 10 of the Algorithm \ref{alg:4.01} are the training from normal data $\boldsymbol{X}^n$, while the steps between 11 and 16 aim the anomaly detection on new contaminated observations. Note that the steps 11 and 14 refers to the md-RPCA, while the steps 12 and 15 refers to the sd-RPCA, and the steps 13 and 16 refers to the kd-RPCA. It is also important to observe that the anomaly detection from new observations does not require new robust subspace learning, it is only necessary to compute a moment-based Mahalanobis distance to classify the $c$ largest distances as anomalous.

However, m-RPCA can also be adopted as unsupervised algorithm through changing the lines 1 and 2 of the Algorithm \ref{alg:4.01} to substitute $\boldsymbol{X}^n$ by $\boldsymbol{X}^t$. Hence, the robust subspace is learned from contaminated data and used for comparing the distance between higher moments  from robust estimate and moments of new observations.

@@

This section presents the performed experiments and the acquired results. First, in Section \ref{sec:4_AnalyzedScenario}, the experimental scenario adopted in the experiments is summarized. Section \ref{sec:4_results} shows the results of the proposed approach, compares to selected anomaly detection algorithms, and presents a discussion about the results.

\subsection{Experimental Scenario}
\label{sec:4_AnalyzedScenario}

Anomaly detection algorithms usually rely on supervised or unsupervised methods, where the former requires labeled normal and anomalous data for training anomaly detection models, while the latter does not require labeled data or training. However, the availability of labeled data is a challenging concern in real world problems of anomalous detection, where anomalies are rare or even unkown events.

Semi-supervised algorithms are an alternative for the anomaly detection problem, considering that this method only relies on normal data for training and that non-malicious data can be obtained from historical information and from rule-based approaches. The proposed approach is semi-supervised and relies on normal data $\boldsymbol{X}^n$ and on contamination rate $c$ for training and anomaly detection, respectively.

Only for evaluation purpose, we use part of the contaminated data, denoted as $\textbf{X}^v$, for cross-validation, in order to estimate the contamination rate $c$ that best detect anomalies from $\textbf{X}^v$. However, we assume that the $c$ is well known or can be estimated for real world problems of anomaly detection.

Clustering methods for anomaly detection rely on three key assumptions: legitimate data instances often fall into a cluster whereas attacks do not, legitimate data instances are usually located near the closest cluster centroid while anomaly ones are often far away from it, legitimate data instances fall into vast and dense clusters and anomalies into small or spare ones \cite{ahmed2016survey, moustafa2019holistic}. Gaussian Mixture Model (GMM) is a statistical-based and parametric approach algorithm that estimates the distribution of the normal class from a training set and is typically based on a set of kernels \cite{moustafa2019holistic}.

This experimental evaluation compares our proposal to widely adopted algorithms for anomaly detection based on clustering and statistical approaches. As clustering-based method, we compare to K-means, which is the clustering algorithm most adopted for general purposes and for network anomaly detection \cite{gaddam2007kmeans,moustafa2019holistic}. It is observed that network traffic do not belong to a Gaussian distribution \cite{benson2010network,moustafa2019holistic}, and that algorithms based on non-Gaussian distributions, such as GMM, has been adopted for network anomaly detection \cite{moustafa2019holistic}. Therefore, we compare our proposal to GMM. We also evaluate the results of ROBPCA for anomaly detection on CTU-13, considering that ROBPCA also relies on robust estimates with adjusted outlyingness based on robust skewness \cite{hubert2009robustskewed}.

>>>>

In this paper, scalars are denoted by italic letters (\emph{a, b, A, B, $\alpha$, $\beta$}), vectors by lowercase bold letters (\textbf{a, b}), matrices by uppercase bold letters (\textbf{A, B}), and $a_i,_j$ denotes the (\emph{i, j}) elements of the matrix \textbf{A}. The superscripts \textsuperscript{T} and \textsuperscript{-1} are used for matrix transposition and matrix inversion, respectively. The Frobenius norm is denoted as $\left\| \mathord{\cdot} \right\|_F$, while $\left\| \mathord{\cdot} \right\|_*$ denotes the nuclear norm of a matrix and $\left\| \mathord{\cdot} \right\|_1$ means the sum of the absolute values of matrix entries. We also define the operator $\langle \mathord{\cdot} \rangle$, that is the standard trace inner product, and $[ \mathord{\cdot} ]^s$, which denotes the $n$ largest values of a vector.

@@

In order to analyse the null hypotheses $H_{1rob.sub}$ and $H_{0rob.sub}$, which evaluate if the robust subspace learning can improves the anomaly detection from imbalanced and skewed data, we create two synthetic data sets characterized by skewed and heavy tailed distributions. We selected Pareto and Lognormal distributions to evaluate the anomaly detection capacity of our proposal, considering that these distributions are skewed and heavy-tailed, and have been adopted to characterize network traffic of the Internet and data centers \cite{benson2010network,leon2017probability}. We also evaluate our proposal and the selected algorithms for anomaly detection on gaussian distributed data set, in order to analyse the detection rate for not skewed and not heavy tailed distributions.

The addictive anomalies for Pareto and Lognormal distributions are gaussian distributed, while an uniform noise is used to add anomalies into the gaussian distributed data set. We generate $\boldsymbol{Y}_p$, $\boldsymbol{Y}_l$ and $\boldsymbol{Y}_g$ for Pareto, Lognormal and Gaussian multivariate distributions, respectively, with 2 features. Each contaminated data set is composed by a number of legitimate observations and the contaminated samples, defined by the adopted $c$.

The Pareto distribution contaminated by gaussian noise is denoted as $\boldsymbol{Y}_p^c$, while $\boldsymbol{Y}_l^c$ denotes the Lognormal distribution contaminated by gaussian noise, and $\boldsymbol{Y}_g^c$ denotes the contaminated gaussian data contaminated by an uniform distribution.

This synthetic data set simulate a total of 2400 events for each scenario, where the number of legitimate observations is simulated according to the desired contamination rate. We evaluate contamination rates $c$ between 1\% and 50\%, to simulate the imbalanced data of anomaly detection problems during our experiments.

The Figure \ref{fig:4.01} depicts the distribution of Pareto and gaussian contamination of one variable, as a visual example of $\boldsymbol{Y}_p^c$, while the Figure \ref{fig:4.02} shows one example of Lognormal contaminated by gaussian noise, generated for $\boldsymbol{Y}_l^c$.

\begin{figure}[!htb]
	\centering
	\subfigure[Pareto and Gaussian Anomalies]{%
		\includegraphics[width=0.47\textwidth]{figures/ch4/4_Xpg2.png}
		\label{fig:4.01}
	}
	\subfigure[Lognormal and Gaussian Anomalies]{%
		\includegraphics[width=0.47\textwidth]{figures/ch4/4_Xlogng2.png}
		\label{fig:4.02}
	}
	\caption[Examples of skewed and heavy tailed distributions]{Examples of skewed and heavy tailed distributions}
	\label{fig:4.03}
\end{figure}

@@

The CTU-13 \cite{garcia2014empirical} is a data set of botnet traffic that was captured in the Czech Technical University, by means of a testbed and malware execution in a real network. The CTU-13 data set contains 13 scenarios with network flows of botnet malwares, that are: neris, rbot, virut, menti, sogou, nsys.ay and murlo. The botnet traffic is also classified as attack or command and control (C\&C), while the legitimate flows can also be classified as normal or background. It is important to note that the CTU-13 data set can also be formulated as a superposition of signal, artefact and noise, which refer to background, normal a botnet traffic, respectively.

@@

We refer to Garcia \cite{garcia2014identifying} and Garcia \emph{et al.} \cite{garcia2014empirical} for a detailed description of the performed attacks and C\&C flows, as well as for more information about the topology of the adopted testbed, rules for classifying normal flows, and an analysis of behaviors or patterns of the malware's traffic.

For all the scenarios, the authors of the CTU-13 data set convert the captured pcap files to NetFlows and release the processed flows. The data set provides ground-truth labels for flows as follows: flows from or to the infected machines are labeled as “botnet”; flows from or to well known and controlled machines are labeled as “normal”; all other flows are labeled as “background.”

Table \ref{tab:4.01} presents an overview grouped by scenario, according to the column ID, and shows the malwares used for botnet attacks, the types of attacks or C\&C, the total number of flows, the number of malicious flows which includes flows of C\&C and attacks, and finally shows the number of normal flows.

@@

The full data set and scenarios can be denoted as $\boldsymbol{X} = \{\boldsymbol{X}_{10}, \boldsymbol{X}_{11}, \ldots , \boldsymbol{X}_{18-2}, \boldsymbol{X}_{19}\}$, in accordance to IDs presented in Table \ref{tab:4.01}. In our experiment each contaminated scenario $\boldsymbol{X}_i$ is split into $\boldsymbol{X}_i^s$ containing 50\% of the the normal data, and into $\boldsymbol{X}_i^c$ that is composed by all anomalous flows and the necessary number of normal flows to have a testing data with the desired contamination rate.

The CTU-13 data set originally contains the following features for each flow:

\begin{itemize}
	\item Start Time;
	\item End Time;
	\item Duration;
	\item Source IP Address;
	\item Source Port;
	\item Direction;
	\item Destination IP Address;
	\item Destination Port;
	\item State of TCP flags;
	\item Destination Type of Service;
	\item Source Type of Service;
	\item Total number of Packets;
	\item Total number of Bytes;
\end{itemize}

Our analysis of the available features leads to discard some features, considering that highly correlated features can bias or not improve the model, and that source or destination IP addresses can insert some false bias into learning models, since they can be changed by IP spoofing. Other risk related to adopt IP address for training models is the training model to learn that
one IP is legitimate and this IP be infected subsequently, which can result into false negative classifications.

We conducted and Exploratory Data Analysis (EDA) on the CTU-13 and observed that some features are skewed and present an overlapping between normal and anomalous flows, as can be seen in Figure \ref{fig:4.05} and Figure \ref{fig:4.05}, that present the distributions of TCP state of scenario 16 and the type of services from destination of scenario 10.

\begin{figure}[!htb]
	\centering
	\subfigure[State of scenario 16.]{%
		\includegraphics[width=0.47\textwidth]{figures/ch4/raw_distplot_capture20110816_State.png}
		\label{fig:4.04}
	}
	\subfigure[Destination Type of service of scenario 10.]{%
		\includegraphics[width=0.47\textwidth]{figures/ch4/raw_distplot_capture20110810_dTos.png}
		\label{fig:4.05}
	}
	\caption[Skewness and Overlapping]{Example of skewness and overlapping}
	\label{fig:4.06}
\end{figure}

However, it is not possible to observe a pattern on distributions of all the features, as depicted by Figures \ref{fig:4.07} and \ref{fig:4.08}, that show the distributions of TCP states of the scenario 10 and source ports of scenario 16, and highlight the distributions of normal and anomalous flows.

\begin{figure}[!htb]
	\centering
	\subfigure[State of scenario 10.]{%
		\includegraphics[width=0.47\textwidth]{figures/ch4/raw_distplot_capture20110810_State.png}
		\label{fig:4.07}
	}
	\subfigure[Source Port of scenario 16.]{%
		\includegraphics[width=0.47\textwidth]{figures/ch4/raw_distplot_capture20110816_Sport.png}
		\label{fig:4.08}
	}
	\caption[Skewness and Overlapping]{Skewness and Overlapping}
	\label{fig:4.09}
\end{figure}

Due to the number of available features and the class overlapping between normal and anomalous flows, we performed an correlation analysis and cross validation in order to identify the best set of features for network attack detection. Therefore, we adopt the following features: state, destination type of service, destination port, source port, total number of packets, total number of bytes and number of bytes from the source.

@@

Robust subspace learning can be defined as the decomposition of a given data matrix $\textbf{X}$ into the sum of a low rank matrix $\textbf{L}$, whose column subspace gives the principal components, and a sparse matrix $\textbf{S}$, with outliers or noise. We propose to learn a robust subspace $\textbf{L}$ from the normal traffic $\textbf{X}^s$ for computing $\textbf{L}^s$, $\textbf{S}^s$ and the robust moments, i.e. the mean $\boldsymbol{\bar{x}}$, skewness $\boldsymbol{\bar{s}}$ and kurtosis $\boldsymbol{\bar{k}}$, in order to evaluate the distance $\boldsymbol{d}$ of new observations $\boldsymbol{X}^c$ to these moments and classify as anomalous the observations with the largest distances.

@@

Before ALM, some methods were proposed to solve that convex optimization problem, such as Iterative Thresholding (IT) and Accelerated Proximal Gradient (APG). However, according to Zhouchen \emph{et al.} \cite{lin2010augmented}, both approaches have scalability problems and require a large number of iterations to converge. The Augmented Lagrange Multiplier is proven to have a \emph{Q-linear} convergence speed and experimental results show that ALM is five times faster than APG, which in theory is sub-linear \cite{lin2010augmented}. Furthermore, ALM reaches more accurate results with less iterations.

@@

We propose to compute the mean $\boldsymbol{\bar{x}}$, the skewness $\boldsymbol{\bar{s}}$, the kurtosis $\boldsymbol{\bar{k}}$ and the covariance matrix $\boldsymbol{\bar{\Sigma}}$ from the robust subspace $\boldsymbol{L}^s$, after to minimize the Equation \ref{eq:4.06} for $\textbf{X}^s$. We also propose to compute the Mahalanobis Distance (MD) for detecting anomalies between new observations and the moments calculated from a robust subspace computed by RPCA.

@@

The contamination rate parameter is widely adopted for well established outlier detection algorithms \cite{zhao2019pyod}, and refers to the percentage of observations that are classified as anomalous. The contamination rate can be well known for some areas, or can be computed by cross-validation or can be assumed according to previous observations. In our proposal, the contamination define the rate of the largest distances $\boldsymbol{d}$ that shall be classified as anomalous. Therefore, the resulting $\boldsymbol{\bar{t}}$ refers to observations classified as anomalous according to the contamination $c$.

The Algorithm \ref{alg:4.01} presents how the m-RPCA can be algorithmically described.
\begin{algorithm}
	\label{alg:4.01}
	\SetAlgoLined
	\KwResult{$\boldsymbol{\bar{t}}_{\bar{\boldsymbol{x}}}$, $\boldsymbol{\bar{t}}_{\bar{\boldsymbol{s}}}$, $\boldsymbol{\bar{t}}_{\bar{\boldsymbol{k}}}$}
	Given $\boldsymbol{X}$ split into $\boldsymbol{X}^s$ and $\boldsymbol{X}^c$\;
	\While{not $\min_{L,S}\left \| \boldsymbol{L} \right \|_{*} + \lambda \left \| \boldsymbol{S} \right \|_{1}$ of $\boldsymbol{X}^s$}{
		$\boldsymbol{L}_{k+1} \in \operatorname*{argmin}_{\boldsymbol{L} \in \mathbb{R}^{m \times n}}\{l(\boldsymbol{L}, \boldsymbol{S}_{k}, \boldsymbol{Y}_{k})\}$\;
		$\boldsymbol{S}_{k+1} \in \operatorname*{argmin}_{\boldsymbol{S} \in \mathbb{R}^{m \times n}}\{l(\boldsymbol{L}_{k+1}, \boldsymbol{S}, \boldsymbol{Y}_{k})\}$\;
		$\boldsymbol{Y}_{k+1} = \boldsymbol{Y}_{k} + \mu(\boldsymbol{X} - \boldsymbol{L}_{k} - \boldsymbol{S}_{k})$\;
	}
	$\bar{\boldsymbol{x}} = \displaystyle\frac{1}{h}\displaystyle\sum_{i\in \boldsymbol{h}} \boldsymbol{x}_i$, for $\boldsymbol{L}^s$\;
	$\boldsymbol{\bar{\Sigma}} = \displaystyle\frac{1}{h}\displaystyle\sum_{i\in \boldsymbol{h}} (\boldsymbol{x}_i - \bar{\boldsymbol{x}})(\boldsymbol{x}_i - \bar{\boldsymbol{x}})^\prime$, for $\boldsymbol{L}^s$\;
	$\boldsymbol{\bar{s}} = \frac{m_3}{m_2^{\frac{3}{2}}}$, for $\boldsymbol{L}^s$\;
	$\boldsymbol{\bar{k}} = \frac{m_4}{m_2^2}$, for $\boldsymbol{L}^s$\;
	$\boldsymbol{d}(\boldsymbol{x},\bar{\boldsymbol{x}}, \boldsymbol{\bar{\Sigma}}) = \sqrt{(\boldsymbol{x} - \bar{\boldsymbol{x}}) \boldsymbol{\bar{\Sigma}}^{-1}(\boldsymbol{x} - \bar{\boldsymbol{x}})^\prime}$\;
	$\boldsymbol{d}(\boldsymbol{x}, \bar{\boldsymbol{s}}, \boldsymbol{\bar{\Sigma}}) = \sqrt{(\boldsymbol{x} - \bar{\boldsymbol{s}}) \boldsymbol{\bar{\Sigma}}^{-1}(\boldsymbol{x} - \bar{\boldsymbol{s}})^\prime}$\;
	$\boldsymbol{d}(\boldsymbol{x}, \bar{\boldsymbol{k}}, \boldsymbol{\bar{\Sigma}}) = \sqrt{(\boldsymbol{x} - \bar{\boldsymbol{k}}) \boldsymbol{\bar{\Sigma}}^{-1}(\boldsymbol{x} - \bar{\boldsymbol{k}})^\prime}$\;
	$\boldsymbol{\bar{t}}_{\bar{\boldsymbol{x}}} = [\boldsymbol{d}(\boldsymbol{x}, \bar{\boldsymbol{x}}, \boldsymbol{\bar{\Sigma}})]^c$\;
	$\boldsymbol{\bar{t}}_{\bar{\boldsymbol{s}}} = [\boldsymbol{d}(\boldsymbol{x}, \bar{\boldsymbol{s}}, \boldsymbol{\bar{\Sigma}})]^c$\;
	$\boldsymbol{\bar{t}}_{\bar{\boldsymbol{k}}} = [\boldsymbol{d}(\boldsymbol{x}, \bar{\boldsymbol{k}}, \boldsymbol{\bar{\Sigma}})]^c$\;
	\caption{Moment Distances from Robust Subspace}
\end{algorithm}

The steps between 1 and 10 of the Algorithm \ref{alg:4.01} are the training from normal data $\boldsymbol{X}^s$, while the steps between 11 and 16 aim the anomaly detection from new contaminated observations. Note that the steps 11 and 14 refer to the md-RPCA, while the steps 12 and 15 refer to the sd-RPCA, and the steps 13 and 16 refer to the kd-RPCA. It is also important to observe that the anomaly detection from new observations does not require new robust subspace learning, it is only necessary to compute a moment-based Mahalanobis distance to classify the $c$ largest distances as anomalous.

The m-RPCA can also be adopted as unsupervised algorithm if the lines 1 and 2 of the Algorithm \ref{alg:4.01} are changed to substitute $\boldsymbol{X}^s$ by $\boldsymbol{X}^c$. Hence, the robust subspace is learned from contaminated data and used for comparing the distance between higher moments from robust estimate and moments of the observations of the same data. It is important to note that this unsupervised approach requires the computational cost of the subspace learning for any new observation.

@@

This section presents the performed experiments on simulated and real data set for anomaly detection. First, in Section \ref{sec:4_metric} we present the adopted metric to evaluate imbalanced data in the context of anomaly detection. The Section \ref{sec:4_SimulatedScenario} describes the experiment for anomaly detection on simulated skewed and heavy-tailed data set, and the Section \ref{sec:4_CTU-13} presents the experiment for network anomaly detection on CTU-13 data set.

\subsection{The metric}
\label{sec:4_metric}

In anomalous detection problems, where anomalies are rare events, if one classify all observations as normal and apply an accuracy evaluation, one would have high accuracy but poor true-positive detection. Due to the importance given by the F-measure to true-positive detection in scenarios such as the network attack detection, it is the preferable measure for imbalanced data sets \cite{powers2011evaluation,moustafa2019holistic}. Therefore, F1-measure is the metric used for validation of our experiments.

This experimental evaluation compares our proposal to widely adopted algorithms for anomaly detection that also rely on contamination rate for anomaly detection, which are: PCA based on the sum of weighted projected distances to the eigenvector hyperplanes \cite{shyu2003novel}; MCD \cite{rousseeuw1984mcd,rousseeuw1999fastmcd}; One-Class Support Vector Machines \cite{scholkopf2001estimating}; Local Outlier Factor (LOF) \cite{breunig2000lof}; k-Nearest Neighbors \cite{angiulli2002fast}; and Isolation Forest \cite{liu2008isolation}.

% ToDO: We also evaluate the results of ROBPCA for anomaly detection on CTU-13, considering that ROBPCA also relies on robust estimates with adjusted outlyingness based on robust skewness \cite{hubert2009robustskewed}.

\subsection{Simulated Experiment}
\label{sec:4_SimulatedScenario}

Anomaly detection algorithms usually rely on supervised or unsupervised methods, where the former requires labeled normal and anomalous data for training anomaly detection models, while the latter does not require labeled data or training.

Semi-supervised algorithms are an alternative for the anomaly detection problem, considering that this method only relies on normal data for training and that non-malicious data can be obtained from historical information and from rule-based approaches. We propose semi-supervised and unsupervised approaches for m-RPCA, where the former relies on normal data $\boldsymbol{X}^s$ for training and on contamination rate $c$ for anomaly detection, while the latter relies only on contaminated data. We assume that the $c$ is well known or can be estimated for real world problems of anomaly detection, in accordance to the selected algorithms for comparison, that also rely on contamination rate.

The availability of labeled data is a challenging concern in real world problems of anomalous detection, where anomalies are rare or even unknown events. Considering that RPCA has already been adopted to isolate outliers from training data \cite{zhou2017anomaly}, we also propose to evaluate the m-RPCA from a semi-supervised approach based on training from contaminated data.

Therefore, we propose to evaluate the following approaches for m-RPCA:
\begin{enumerate}
	\item Semi-supervised;
	\item Contaminated Semi-supervised;
	\item Unsupervised.
\end{enumerate}

For the semi-supervised approach we propose to learn the robust subspace and compute the moments from the normal data set $\boldsymbol{Y}_g$, $\boldsymbol{Y}_p$ and $\boldsymbol{Y}_l$ with Gaussian, Pareto and Lognormal distributions, respectively, and test the anomaly detection for contaminated data set $\boldsymbol{Y}_g^c$, $\boldsymbol{Y}_p^c$ and $\boldsymbol{Y}_l^c$.

In the Contaminated Semi-supervised approach we evaluate the robustness of the m-RPCA for learning from contaminated training data, in order to analyse if m-RPCA can be an alternative for the lack of known normal data. Therefore we propose to train the model from a contaminated normal data $\boldsymbol{Y}_g^{c'}$, $\boldsymbol{Y}_p^{c'}$ and $\boldsymbol{Y}_l^{c'}$, with the same contamination rate of the testing data, but without data sharing between training and testing, taking into account that we shall consider a different contamination data but the same contamination rate.
